Analysing Constraint Grammars with a SAT-solver

### 2016-02-29 paper/lrec/problem_desc.tex
small changes to the intro to SAT

Constraint Grammar (CG, \newcite{karlsson1995constraint})
% \citep[CG,][]{karlsson1995constraint} 
is a formalism used to disambiguate morphologically analysed text. 
A grammar consists of rules that target specific analyses for selection or removal, based on contextual tests. For example, the following rule
\begin{itemize}
\item[] \texttt{REMOVE verb IF (-1 det) ;}
\end{itemize}
removes a verb reading from a word which is preceded by a determiner.
Given the following text,
\begin{itemize}
\item[] 
\begin{verbatim}
"<the>"
        "the" det def
"<wish>"
        "wish" noun sg
        "wish" verb pl
        "wish" verb inf
\end{verbatim}
\end{itemize}
the rule will match to the word \emph{wish}, and remove the analyses \texttt{verb pl} and \texttt{verb inf}.
However, if the target word has only one remaining analysis, then the rule will not apply, even if the condition is met.

CGs are valuable resources for rule-based NLP, especially for lesser
resourced languages. They are robust and can be written without large
corpora---only morphological analysis is needed. The formalism is
lightweight and language-independent, and resources can be shared
between related languages \cite{bick2006spanish,lene_trond_linda2010}.
Mature CGs contain some thousands of rules, but even small CGs are shown to be effective \cite{lene_trond2011}.

@@

However, this design is not without problems. As CGs grow larger, it
gets harder to keep track of all the rules and their interaction.
Our tools will help grammar writers and users to find conflicting
rules, diagnose problems and improve their grammars. 
We expect two major use cases: 
first, to test the effect of new rules while writing a grammar, and
second, to take a complete grammar and analyse it as a whole, to find
conflicts or dead rules.

@@

For the second use case, here are examples of conflicts that our tools will detect.
\begin{itemize}
\item If a rule appears twice, the second occurrence will be disabled by the first
\item $R$ selects something in a context, $R'$ removes~it
\item $R$ removes something from the context of $R'$, so $R'$ can never
  apply
\item $R$ has an internal conflict, such as non-existent
tag combination, or contradictory requirements for a context word
\end{itemize}
$R$ can also be a set of rules: for instance, if one rule removes a verb in
context $C$, and another in context $\neg C$, together these rules
remove a verb in all possible cases, disabling any future rule that
targets verbs.

While rule-internal conflicts can be detected by simpler means, taking
care of rule interaction requires a {\em semantic} rather than a {\em
 syntactic} analysis.
In order to find effects of rule interaction, we must keep track of
the possible sentences at each step. After each rule, we have two
possibilities: the rule fires, or it does not fire. In case the rule does
not fire, we have again two options: either its conditions are not met,
or its target is the only remaining analysis.

We express these requirements as a \emph{Boolean satisfiability problem} (SAT).
SAT problems consist of two components: a set of Boolean variables, and a set
of clauses on those variables. For instance, let the set be $\{a,b\}$
and the formulas $\{a \vee b, \neg{}a\}$. A program called \emph{SAT
  solver} will try to find a solution, where all the variables have a
value. For this particular problem, the unique solution is $a=False,
b=True$. It is also possible for a SAT problem to have no solution, or
multiple solutions.

>>>>

Constraint Grammar (CG, \newcite{karlsson1995constraint})
% \citep[CG,][]{karlsson1995constraint} 
is a formalism used to disambiguate morphologically analysed text. 
A grammar consists of rules that target specific readings for selection or removal, based on contextual tests. For example, the following rule
\begin{itemize}
\item[] \texttt{REMOVE verb IF (-1 det) ;}
\end{itemize}
removes all verb readings from a word which is preceded by a determiner.
Given the following text,
\begin{itemize}
\item[] 
\begin{verbatim}
"<the>"
        "the" det def
"<wish>"
        "wish" noun sg
        "wish" verb pl
        "wish" verb inf
\end{verbatim}
\end{itemize}
the rule will match to the word \emph{wish}, and remove the readings
\texttt{"wish" verb pl} and \texttt{"wish" verb inf}.
However, if the target word has only one remaining reading, then the
rule will not apply, even if the condition is met.

CGs are valuable resources for rule-based NLP, especially for lesser
resourced languages. They are robust and can be written without large
corpora---only morphological analyser is needed. The formalism is
lightweight and language-independent, and resources can be shared
between related languages \cite{bick2006spanish,lene_trond_linda2010}.
Mature CGs contain some thousands of rules, but even small CGs are
shown to be effective \cite{lene_trond2011}.

@@

However, this design is not without problems. As CGs grow larger, it
gets harder to keep track of all the rules and their interaction.
Our tool will help grammar writers and users to find conflicting
rules, diagnose problems and improve their grammars. 
We expect two major use cases: 
first, to test the effect of new rules while writing a grammar, and
second, to take a complete grammar and analyse it as a whole, to find
conflicts or dead rules.

@@

For the second use case, here are examples of conflicts that our tool will detect.
\begin{itemize}
\item If a rule appears twice, the second occurrence will be disabled by the first
\item $R$ selects something in a context, $r$ removes~it
\item $R$ removes something from the context of $r$, so $r$ can never
  apply
\item $r$ has an internal conflict, such as non-existent
tag combination, or contradictory requirements for a context word
\end{itemize}
$R$ can be one rule or a set of rules: for instance, if one rule removes a verb in
context $C$, and another in context $\neg C$, together these rules
remove a verb in all possible cases, disabling any future rule that
targets verbs.

While rule-internal conflicts can be detected by simpler means, taking
care of rule interaction requires a {\em semantic} rather than a {\em
 syntactic} analysis.
% In order to find effects of rule interaction, we cannot just apply the
% rules to some actual sentences and hope to find a conflict---corpora
% can be extremely valuable in the development of CG rules, but there
% are languages where corpora are not available, or they may contain
% text from limited domains. 
We must keep track of all the possible sentences after applying each
rule. At each step, we have two options: either the rule fires,
or it does not fire. In case the rule does not fire, we have again two
options: either its conditions are not met,
or its target is the only remaining analysis. This results in a complex table
of interdependent decisions---some determined by being the target of a
rule, others by being a condition.

We express these constraints as a \emph{Boolean satisfiability problem} (SAT).
SAT problems consist of two components: a set of Boolean variables, and a set
of clauses on those variables. For instance, let the set be $\{a,b\}$
and the formulas $\{a \vee b, \neg{}a\}$. A program called \emph{SAT
  solver} will try to find a solution, where all the variables have a
value. For this particular problem, the unique solution is $a=False,
b=True$. It is also possible for a SAT problem to have no solution, or
multiple solutions.
